---
title: "Team Report"
author: "Sam Coleman, Ian Eykamp, Carlos Godinez, Evelyn Kessler, Isabel Serrato"
date: 10/19/2021
output:
  github_document:
    toc: true
---

## Background

In late 2019 a novel coronavirus quickly spread throughout the global population causing a pandemic that is still causing massive disruptions in late 2021. SARS-CoV-2, the virus that causes COVID-19, is a highly contagious respiratory virus. In response to the rapidly spreading disease and it's even more infectious variants, local governments and counties throughout the pandemic have instituted face covering and mask mandates to slow the spread. While specifics change from location to location, a mask mandate largely requires individuals to cover their face and noses when in a public setting. The intent behind this is to reduce the viral load that a potentially infectious person could spread to others [1]. Masks have been found to be effective in slowing the spread of COVID-19 and are recommended to be worn by the CDC when in geographical areas of high spread [2].

## Dataset Used

For this challenge, we used two datasets provided in the Covid challenge, the [population data from the US census](https://data.census.gov/cedsci/table?t=Population%20Total&g=0100000US%240500000&tid=DECENNIALPL2020.P1), as well as Covid cases and deaths by county over time taken from a [dataset collected by the New York Times](https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv). The population data is from the US Census Bureau and is collected from the American Community Survey (ACS) Total Population by County 5 year estimate. An important note about this data: "Estimates of urban and rural populations, housing units, and characteristics reflect boundaries of urban areas defined based on Census 2010 data. As a result, data for urban and rural areas from the ACS do not necessarily reflect the results of ongoing urbanization." We did not look into how our results would change if there was a higher population in urban areas, however, it would be interesting to re-do our analysis once the most recent 2020 census data is released and see if there are any significant changes. While the population data is the most accurate data we can obtain currently, it isn't perfect and is a bit out of date, both of which are limiting factors of this dataset. The COVID-19 data is obtained from the New York times and includes cumulative cases and deaths over time for each county from January 21, 2020 to October 17, 2021. The dataset obtains the time series data "from state and local governments and health departments in an attempt to provide a complete record of the ongoing outbreak." The documentation for the table provides an important note about potential limitations: "Since the first reported coronavirus case in Washington State on Jan. 21, 2020, The Times has tracked cases of coronavirus in real time as they were identified after testing. Because of the widespread shortage of testing, however, the data is necessarily limited in the picture it presents of the outbreak." We were not concerned about this limitation, as this is a problem every dataset about COVID-19 will encounter since not everyone who has had COVID-19 has received a test for a variety of reasons.

We are using a [Mask Mandate CDC database](https://data.cdc.gov/Policy-Surveillance/U-S-State-and-Territorial-Public-Mask-Mandates-Fro/62d6-pm5i) to determine whether or not a mask mandate was in place in every county for every day from April 10, 2020 through August 15, 2021. These data are "derived from publicly available state and territorial executive orders, administrative orders, resolutions, and proclamations ("orders") for COVID-19 that expressly require individuals to wear masks in public found by the CDC, COVID-19 Community Intervention & Critical Populations Task Force, Monitoring & Evaluation Team, Mitigation Policy Analysis Unit, Center for State, Tribal, Local, and Territorial Support, Public Health Law Program, and Max Gakh, Assistant Professor, School of Public Health, University of Nevada, Las Vegas." Two other important notes are that "any orders not available through publicly accessible websites are not included in these data" and that "These data do not include data on counties that have opted out of their state mask mandate pursuant to state law." Therefore, for each date, every county either has a "Yes" (there was a mask mandate in place), "No" (there was not a mask mandate in place), or NA (it is unknown whether or not there was a mask mandate in place).

We identified two main limitations of this data set. The first is the amount of NAs in the dataset, and the difficulty this presented in knowing how to best complete our analysis with the lack of complete data. If the data was not publicly accessible by the entities mentioned above, an NA was placed in the mask mandate column. As a quick check, some members of our team looked at their home counties and saw NAs during times they knew there was a mask mandate in place, such as counties in California. There are other states, such as Florida, where there are only NAs for every county through the entire dataset. We still chose to analyze this dataset as after conducting further research, it is the most complete mask mandate dataset that was available. It is difficult to have a perfect dataset with information about mask mandates without having central database for which counties individually report to. In our analysis, for counties where there was at least one day with a yes/no mask mandate, we treat the NAs as "No" values.

The second limitation of the dataset is that a mask mandate does not necessarily equal compliance. We thought about different ways we could take this into account and looked at a dataset of self-reported mask-wearing survey. We decided not to pursue incorporating this dataset into our analysis as it only covered July 2 to July 14, 2020, a relatively early stage in the pandemic. We did not attempt to minimize this limitation in our analysis, but it is important to consider that mandate does not necessarily equal people wearing masks, or wearing masks correctly.

## Team Question

During our exploratory data analysis (EDA) we looked into normalized cases and deaths vs. self-reported mask wearing, mask mandate trends geographically, and the relationship between normalized cases and whether or not a mask mandate was in place, we arrived at the question: What is the effect of mask mandates on Covid Case Counts per one hundred thousand residents?

## Our Graph

```{r}
library(tidyverse)
load("data/cases_mandate_new_york_fixed.RData")
load("data/mask_deaths")
```

```{r}
us_3 %>%
  distinct(region, .keep_all = TRUE) %>% 
  pivot_longer(
    cols = c(cases_per100k, deaths_per100k),
    names_to = "cases_or_deaths",
    values_to = "per100k"
  ) %>% 
  ggplot(mapping = aes(x = days_with_mandate, y = per100k, color = cases_or_deaths)) +
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_grid(cases_or_deaths ~ ., scales = "free_y", 
             labeller = as_labeller(c(`cases_per100k` = "Cases per 100,000", `deaths_per100k` = "Deaths per 100,000"))) +
  theme_test() +
  theme(
    legend.position = "none",
    plot.title.position = "plot",
    panel.border = element_rect(fill = NA, size = 1),
    # panel.background = element_rect(fill = "grey90"),
    panel.grid.major = element_line(color = "grey50"),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    strip.text.y = element_text(size = 14),
    
  ) +
  labs(
    x = "Cummulative percent of people under mask mandate (%)",
    y = "",
    title = "Mask mandates correlate with Covid-19 cases more strongly than deaths",
    subtitle = "State-level data from Jan 21, 2020 to Oct 17, 2021"
  ) 
```

## Conclusion

We found that there is a negative correlation between the average percentage of days that each county had a mask mandate and the number of Covid cases per 100,000 residents. At the same time, there is not a strong correlation between the average percent of days that each county had a mask mandate and the number of deaths per 100,000 residents. Therefore, the dataset supports the claim that mask mandates may help to reduce the number of cases but not the number of deaths from COVID-19.

## Future Exploration

We would be curious to explore the amount of lag that occurs between when a mandate is put in place and when the rate of increase in cases per day begins decreasing. In other words, quantitatively finding how long it takes for a mask mandate to become effective for curbing the increase in COVID cases in a county. Developing an algorithm to do this would allow us to explore whether there is a difference in the lag depending on any other factors (such as location or politics).

An additional area to explore would be to find a dataset relating whether or not people in a county actually wore masks while a mandate is in place. Our group found a New York Times poll from summer of 2020 that showed how often residents in every county wore masks. However, the outdated nature of this dataset made it difficult to compare with the mandate data. It would be interesting to see the relationship between mask mandates and whether or not residents follow the mandates as it relates to the number of cases and deaths in those counties.

# Notes

<!-- -------------------------------------------------- -->

[1] <https://pubmed.ncbi.nlm.nih.gov/32512240/>

[2] <https://www.cdc.gov/coronavirus/2019-ncov/prevent-getting-sick/cloth-face-cover-guidance.html>
