---
title: "Cross-border Covid Correlation"
author: "Ian Eykamp"
date: "10/17/2021"
output: github_document
---

# Cross-border Covid Correlation

This document and `actualGraph.Rmd` are for visualizing the cross-border correlation data created in the script `neighbors.R`.

Correlations in Covid-19 case rates are computed for each pair of neighboring states in the contiguous US using the following formula. The difference between successive daily totals is calculated using the `diff` function, to give the number of new cases reported on that day. The rows of `state1` and `state2` are joined by date, but the date itself is not used. Then the correlation coefficient is calculated using `cor(state1, state2)`.

The same correlation is then computed when each state is given an offset of between 1 and 100 days, using the `lag` parameter in `diff`. The idea is that states that lag behind one another in their cross-border correlation is an indicator of cross-border flow of Covid-19 cases from the leading state to the lagging state.

Take Alabama and Florida. Their simultaneous correlation of Covid-19 cases during the pandemic has an `r` coefficient of `0.537`. This is not a particularly strong correlation, especially compared with that between other states, and it is more likely to reflect the national trend in cases rather than a direct connection between the states in question.

However, the lagging correlations tell a different story. When the Florida case values are held constant and Alabama's values are given a variable time lag, the maximum correlation between the two states occurs when the Alabama data have a lag of 5 days, yielding a correlation coefficient of `0.624`.

By contrast, when the Alabama case values are held constant, the maximum correlation occurs when Florida's cases have a lag of 17 days (meaning Florida's case numbers are up to 17 days *behind* Alabama's), yielding a correlation of `0.751`.

This asymmetry suggests a directionality to the flow of cases: Alabama's case values changed first, and 2-3 weeks later, Florida saw a very similar change in case numbers. Not only is the time lag for the Alabama-to-Florida direction greater, there is also a much stronger correlation in this direction.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
load("C:/dev/git/covid-project/spatial_correlation/stateCorrs_different_lags.RData")
```

The following plot illustrates the imbalance in the lagged correlations

```{r lagged correlations}
state1 <- "alabama"
state2 <- "florida"

stateCorrs %>%
  semi_join(
    tibble(
      state = c(state1, state2), 
      neighbor = c(state2, state1)
    ), 
    by = c("state" = "state", "neighbor" = "neighbor")
  ) %>% 
  ggplot(mapping = aes(x = lag, y = corr, color = state)) +
  geom_line() + 
  geom_vline(xintercept = c(5, 17), linetype = "dashed", alpha = 0.5) + 
  labs(
    x = "Time lag (days)", 
    y = "Correlation coefficient (r)", 
    color = "State with\nlag applied", 
    title = "Temporal Asymmetry between Alabama and Florida"
  )
```

## Looking for patterns

To see if it is always true that a longer time lag corresponds with a higher maximum correlation, see the scatterplot below. There does not appear to be any correlation. There is a cluster of points with very low correlation and some with unusually long time lags (\> 35 days = 5 weeks), but these do not correspond to the same state.

```{r scatterplot}
df_2 <- 
  stateCorrs %>% 
  group_by(state, neighbor) %>% 
  filter(corr == max(corr)) %>% 
  ungroup() %>% 
  rename(max_lag = lag)

df_2 %>%
  ggplot(mapping = aes(x = max_lag, y = corr, color = state)) +
  geom_point(show.legend = FALSE)
```

The following histogram shows that most correlations between neighboring states are strongest for very small lags (the median is 6 days). This indicates that the trajectory of the pandemic was largely uniform across the US, as half of the states experienced changes in case rates within a week of each other.

```{r histogram}
df_2 %>% 
  summarize(median = median(max_lag)) %>% 
  print()

df_2 %>%
  ggplot(mapping = aes(x = max_lag)) +
  geom_histogram(binwidth = 1)


```

As stated, the correlations are calculated when a lag is applied to each state separately, and this often yields asymmetries in correlation. This last visualization shows the difference in maximum correlation and the corresponding lag that produced it for each pair of neighboring states. For Florida and Alabama, for example, the difference in correlations would be `.751 - .624 = .127`, and the corresponding lag would be `17 - 5 = 12` days. There is an apparent linear correlation between these two quantities! This means that pairs of neighboring states for which cases in one state lagged behind the other are in general more strongly correlated if the lag time between them is greater.

```{r flow}

df_sorted <- 
  df_2 %>% 
  mutate(
    switched = state > neighbor, 
    # if switched is TRUE, it's [neighbor, state], else it's [state, neighbor]
    tmp_state    = if_else(switched, neighbor, state), 
    tmp_neighbor = if_else(switched, state, neighbor), 
    # now replace with the switched values
    state = tmp_state, 
    neighbor = tmp_neighbor
  ) %>% 
  select(!c("tmp_state", "tmp_neighbor")) %>% 
  arrange(state, neighbor)

df_sorted_2 <- 
  df_sorted %>% 
  group_by(state, neighbor) %>% 
  mutate(
    directional_flow = if_else(switched, -max_lag, max_lag), 
    directional_corr = if_else(switched, -corr, corr)
  ) %>% 
  summarize(
    flow = sum(directional_flow), 
    corr = sum(directional_corr), 
  ) %>% 
  ungroup()

df_sorted_2 %>%
  ggplot(mapping = aes(x = flow, y = corr)) +
  geom_point(aes(color = state), show.legend = FALSE) + 
  geom_smooth(method = "lm") + 
  labs(
    x = "Difference in lag at max. correlation (days)", 
    y = "Difference in max. correlation coefficient"
  )

save(
  df_sorted, 
  file = "C:/dev/git/covid-project/spatial_correlation/covid_flows_sorted.RData"
)
```

In `actualGraph.R`, the time lags are shown as directional arrows; states that where cases changed first are shown with arrows pointing to the state with a lagging change. The natural interpretation when seeing that graph (yes, it is an *actual graph* in the mathematical sense) is to see arrows as flows of increasing case numbers. Remember these two points when reading it.

-   Does any of this imply causation -- that *because* the case numbers for one state lagged behind the other, that cases *moved* or *overflowed* from one state to the other? No. But it gives us some insight into the temporal relationship between neighboring states, and where case rate changes happened first.

-   It is also important to note that correlations can be for changing case numbers up, or down -- a state that was ahead of another state could have either seen cases fall first or rise first.
